---
title: "Join Lab"
author: "Tyler Guerrero"
date: "DATA 2401"
output: html_document
---


# Joins and `dplyr` practice

## Dataset 1

We start by making some "healthcare" data: 

```{r}
# Patients table
patients <- tibble(
  patient_id = c(1, 2, 3, 4),
  name = c("Alice", "Bob", "Charlie", "David"),
  age = c(25, 40, 35, 50),
  sex = c("F", "M", "F", "M")
)

# Visits table (Some patients have multiple visits, some missing)
visits <- tibble(
  visit_id = c(101, 102, 103, 104, 105),
  patient_id = c(1, 2, 2, 3, 5),  
  doctor_id = c(201, 202, 201, 203, 204),
  visit_date = as.Date(c("2024-01-10", "2024-01-12", "2024-02-15", "2024-03-01", "2024-03-10"))
)

# Doctors table (Some doctors may not have treated any patients)
doctors <- tibble(
  doctor_id = c(201, 202, 203, 205),
  doctor_name = c("Dr. Smith", "Dr. Johnson", "Dr. Lee", "Dr. Brown"),
  specialty = c("Cardiology", "Neurology", "Pediatrics", "Orthopedics")
)

```

Questions: 

- Find the names of patients who have visited the hospital. 
- Show all patients visit information, even if they haven't visited the hospital
- List all the visits and patient information, even if there isn't information for the patient
- Show all visits with their assigned doctors.
- Find the doctor(s) that Bob has met with. 
- Find patients that haven't visited the hospital

```{r}
## Names of patients who have visited hospital.
inner_join(patients, visits)

## All patients visit information.
left_join(patients, visits)

## All visits and patient information, even if there isn't patient information.
left_join(visits, patients)

## Show all visits with assigned doctors.
left_join(visits, doctors)

## Doctors Bob has met with.
left_join(patients, visits) %>% left_join(doctors) %>% filter(name == "Bob")

## Patients that haven't visited the hospital.
anti_join(patients, visits)
```

## Dataset 2

Load the `"nycflights13"` package.  You'll also need to load `dplyr` or `tidyverse`, if you didn't above.

```{r}
library(tidyverse)
#install.packages("nycflights13")
library(nycflights13)
my_flights <- flights # store my own version
my_flights <- na.omit(my_flights) ## this dataset has data on cancelled flights too, 
                                  ## so we 'll remove those for now

```


Create a dataframe of the average arrival delays for each _destination_

```{r}
delay_by_dest <- my_flights %>% group_by(dest) %>% summarise(average_arrival_delay = mean(arr_delay))

```

Then use `left_join()` to join on the "airports" dataframe, which has the airport information.

```{r}
glimpse(airports)
glimpse(delay_by_dest)

delay_by_dest %>% 
  inner_join(airports, join_by(dest == faa)) %>% 
  select(name, average_arrival_delay) %>% 
  arrange(-average_arrival_delay)
  
```


Note here: the column names don't match anymore! 

Which airport had the largest average arrival delay?

```{r}
delay_by_dest %>% 
  inner_join(airports, join_by(dest == faa)) %>% 
  select(name, average_arrival_delay) %>% 
  slice_max(average_arrival_delay)

## Columbia Metropolitan: 41.76415
```



What is the average departure delay for each airline? To answer this with something more than just the FAA code of the airline name, you need to create a dataframe of the average departure delay for each _airline_, then use `left_join()` to join on the "airlines" dataframe


```{r}
glimpse(airlines)

my_flights %>% 
  group_by(carrier) %>% 
  summarise(average_dep_delay = mean(dep_delay)) %>% 
  left_join(airlines, join_by(carrier)) %>% 
  select(name, average_dep_delay) %>% 
  arrange(-average_dep_delay)


```

Which airline had the smallest average departure delay? 



```{r}

my_flights %>% 
  group_by(carrier) %>% 
  summarise(average_dep_delay = mean(dep_delay)) %>% 
  left_join(airlines, join_by(carrier)) %>% 
  select(name, average_dep_delay) %>% 
  slice_min(average_dep_delay)

## US Airways Inc.
```


Add origin and destination airport names (two joins to the same table)
Question: Which origin–destination pair has the highest average arrival delay?


```{r}



``` 

Which airline (name) has the highest on-time rate? Are left_join and inner_join returning the same rows here? Why?


```{r}



```

Plane metadata join: age of aircraft vs. delays
Among aircraft with at least 50 flights, which tailnumber has the largest average arrival delay, and what are the manufacturer/model and age?


```{r}



```

## Extra Examples from R4DS:    

```{r}

top_dest <- flights %>%
  count(dest, sort = TRUE) %>%
  head(10)
top_dest

```

Now you want to find each flight that went to one of those destinations. You could, of course, filter. But! Now you can semi-join: 

```{r}
flights %>% 
  semi_join(top_dest)
```


What does anti_join(flights, airports, by = c("dest" = "faa")) tell you? What does anti_join(airports, flights, by = c("faa" = "dest")) tell you?
Perform the two joins below and consider what the results **are**. 

```{r}

```

